One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-12 paper/model.tex
minor updates

\section{Governing equations and numerical models}\label{model} 
We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
orbting a central star of mass $M_*$. We will mainly examine two-dimensional
(2D, or razor-thin) discs in favour of numerical 
resolution, but have also carried out some three-dimensional (3D) disc
simulations. Thus, for generality we describe the system in
3D, using both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}=(v_r,v_\theta,v_\phi)$ the
fluid velocity, $p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed is given by 
\begin{align}\label{sound-speed}
  % c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio at the reference radius $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient (since, for an ideal gas the
temperature is proportional to $c_s^2$). The vertical disc scale-height is
defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
a disc with constant $H/R$.

For the simulations presented in this paper, the indirect potential
associated with the non-inertial reference frame is purposefully 
neglected to avoid complications arising from the motion of the
central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
effect is not expected to be significant \citep{adams89,shu90}. Indeed,
simulations in the early stages of this project included the indirect
potential, and yield similar results.

% ang mom cons
% The locally isothermal
% equation of state allows us to apply standard linear density wave
% theory for results interpretation. 
% This allows us to consider
%an isolated disc.

The 2D disc equations are obtained by replacing $\rho$ with the surface
mass density $\Sigma$, $p$ becomes the vertically-integrated pressure, 
and the 2D fluid
velocity $\bm{v}=(v_R,v_\phi)$ is evaluated at the midplane, as are
the forces from the potential. In the Poisson equation, $\rho$ is
replaced by $\Sigma\delta(z)$, where $\delta(z)$ is the
delta function.

\subsection{Disc model and initial conditions}
For the initial disc profile we adopt a modified power-law disc with
surface density given by
\begin{align}
  \Sigma(R) = \Sigma_\mathrm{ref} \left(\frac{R}{R_0}\right)^{-s}\times B(R;
  R_{1}, R_{2}, \epsilon, \delta R), 
\end{align}
where $s$ is the power-law index describing the smooth disc and
$\Sigma_\mathrm{ref}$ is a 
surface density scale chosen by specifying $Q_\mathrm{out}$,
the Keplerian Toomre parameter at $R=R_{2}$,
\begin{align}
  Q_\mathrm{out} = \left.\frac{c_s\Omega_k}{\pi G
    \Sigma}\right|_{R=R_{2}}. 
\end{align}
The bump function
$B(R)$ represents a surface density boost between
$R\in[R_{1},R_{2}]$ by a factor $\epsilon^{-1}>1$,
and $\delta R$ is the transition width between the bump and the
smooth disc. We choose 
\begin{align}
  &B(R) = f_1(R)\times f_2(R),\\
  &f_1(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 +
    \tanh\left(\frac{R-R_{1}}{\Delta_1}\right)\right]  + \epsilon,\\
  &f_2(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 -
    \tanh\left(\frac{R-R_{2}}{\Delta_2}\right)\right]  + \epsilon,
\end{align}
where $\Delta_1 = \delta R \times H(R_{1})$ and similarly for
$R_{2}$. The bump function mimicks mass built-up at some location in 
the disc.

@@

Our fiducial parameter values are: $s=2$, $R_{1}=R_0$, $R_{2}=2R_0$,
$\epsilon=0.1$, $\delta R=5$, $h=0.05$ and
$Q_\mathrm{out}=2$. An example of the initial surface density and the
Toomre parameter (see\S\ref{wkb}) is shown in
Fig. \ref{initial_surf}. The transition from the self-gravitating and
non-self-gravitating portions of the disc occur smoothly across
$\sim10H$. Initially there is no vertical or radial velocity
($v_R = v_r = v_\theta = 0$). The azimuthal velocity is initialized to
satisfy centrifugal balance with pressure and gravity,
\begin{align}
  \frac{v_\phi^2}{r} = \frac{1}{\rho}\frac{\p p}{\p r} + \frac{\p
    \Phi_\mathrm{tot}}{\p r}
\end{align}
and similarly in 2D. The angular velocity is $\Omega \equiv v_\phi/R$.

@@

\subsection{FARGO}
Our primary tool is FARGO with self-gravity \citep{baruteau08}. This
is a popular, simple finite-difference code for 2D discs. `FARGO' refers
to its azimuthal transport algorithm, which circumvents time-step
contraint set by the inner disc boundary \citep{masset00a,masset00b}.

The 2D disc occupies
$R\in[R_\mathrm{min},R_\mathrm{max}],\,\phi\in[0,2\pi]$ and is 
divided into $(N_R,N_\phi)$ grids, logarithimically spaced in radius and
uniform spaced in azimuth. At radial boundaries we set the
hydrodynamic variables to their initial values.

@@

The normalised surface density associated with $m$ is
\begin{align}
  \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
      m\phi)}\right]
\end{align}
where $\Sigma_{00} = \Sigma_0(t=0)$. The time evolution of the
$m^\mathrm{th}$ mode can be characterized by 
assuming $\Sigma_m\propto\exp{(-\ii \sigma t)}$, where
$\sigma = \omega + \ii\gamma$ is the complex frequency, $\omega$ is
the real frequency and $\gamma$ is the growth rate. The total
non-axisymmetric surface density is 
\begin{align}
  \Delta\Sigma = \frac{\Sigma - \Sigma_0}{\Sigma_0}. 
\end{align}

\subsubsection{Angular momentum decomposition}
The total disc angular momentum is
\begin{align}
  J &= \int_{R_\mathrm{min}}^{R_\mathrm{max}}\int_0^{2\pi} \Sigma Rv_\phi RdRd\phi \notag\\
  &= 2\pi\int_{R_\mathrm{min}}^{R_\mathrm{max}} R\Sigma_0 v_{\phi0} R dR \notag\\ 
  &\phantom{=}+
  \sum_{m=1}^\infty2\pi\int_{R_\mathrm{min}}^{R_\mathrm{max}} 2R\real\left[\Sigma_m v_{\phi
      m}^*\right] RdR 
  %&= \sum_{m=0}^\infty \int _{R_\mathrm{min}}^{R_\mathrm{max}}  j_m(R)
  %dR 
  = \sum_{m=0}^\infty J_m. 
\end{align}
where $^*$ denotes complex conjugation. We shall refer to $J_m$ as the
$m^\mathrm{th}$ component of the total angular momentum, and use it to
monitor angular momentum conservation in the numerical simulations. 
It is important to distinguish this empirical definition from the
angular momentum of linear perturbations, defined through a
conservation law, which is given in \S\ref{wkb}.

@@

\subsubsection{Three-dimensionality}
In 3D simulations we measure the importance of vertical motion with
$\Theta$, where 
\begin{align}\label{theta}
  \Theta^2 \equiv \frac{\avg{v_z^2}}{\avg{v_R^2}+\avg{v_\phi^2}}, 
\end{align}
and $\avg{\cdot}$ denotes the density-weighted average, e.g., 
\begin{align}
  \avg{v_z^2} \equiv\frac{
    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
    \rho v_z^2 dV}{
    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
    \rho dV
  },
\end{align}
and similarly for the horizontal velocities. The radial range of
integration is taken over $r\in[R_1,R_2]$ since this is where
we find the perturbations to be confined.

\section{Linear density waves}\label{wkb}
In this paper we take a fully numerical approach using non-linear
hydrodynamic simulations. However, it will be useful to employ
existing linear density wave theory to interpret the simulations.
We list here the main definitions and results from 2D linear theory,
taken from \cite{lin11b} and \cite{shu91}.

@@

The background torque density $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state. In a barotropic fluid, such as a strictly isothermal disc,
$\tbg$ vanishes and the angular momentum associated with
the perturbation is conserved, provided that there is no net angular
momeumtum flux.

@@

\subsection{Local results}
In the local approximation, perturbations are assumed to vary rapidly
relative to any background gradients. \cite{shu91} gives the
dispersion relation for tightly-wound density 
waves of the form $\exp{[\ii(-\sigma t + m \phi + kR)]}$ in a razor-thin
disc as  
\begin{align}\label{dispersion}
  (\sigma - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
\end{align}
and $k$ is a real wavenumber such that $|kR|\gg1$. Note that in this
approximation, only axisymmetric perturbations with $m=0$ can be
unstable (having an imaginary $\sigma$).

Considering a neutral mode, given the real frequency $\omega$ or
pattern speed $\Omega_p\equiv 
\omega/m$, Eq. \ref{dispersion} can be solved 
for $|k|$, 
\begin{align}\label{wavenumber}
  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
      Q^2(1-\nu^2)}\right], 
\end{align}
where 
% \begin{align}
%   k_T \equiv \frac{\kappa^2}{2\pi G \Sigma}
% \end{align}
% is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. In
Eq. \ref{wavenumber}, the upper (lower) sign correspond to short
(long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
spirals.

@@

A detailed discussion of the properties of density waves in the local
approximation can be found in \cite{shu91}. An important result is that
waves interior to co-rotation ($R<R_c$) have negative angular momentum, while
waves outside co-rotation ($R>R_c$) have positive angular
momentum. The local approximation becomes less good for gobal modes,
but will nevertheless be useful for results interpretation.

>>>>

\section{Governing equations and numerical models}\label{model} 
We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
orbting a central star of mass $M_*$. We will mainly examine 2D
(or razor-thin) discs in favour of numerical 
resolution, but have also carried out some 3D disc
simulations. Thus, for generality we describe the system in
3D, using both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the 
fluid velocity (the angular velocity being $\Omega \equiv v_\phi/R$), 
$p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed $c_s$ is given by 
\begin{align}\label{sound-speed}
  % c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio at the reference radius $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient (since, for an ideal gas the
temperature is proportional to $c_s^2$). The vertical disc scale-height is
defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
a disc with constant $H/R$.

For the simulations presented in this paper, the indirect potential
associated with the non-inertial reference frame is purposefully 
neglected to avoid complications arising from the motion of the
central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
effect is not expected to be significant
\citep{adams89,shu90}. Indeed, simulations including the indirect
potential, carried out in the early stages of this project, produced
similar results.

%isolated system 
% ang mom cons
% The locally isothermal
% equation of state allows us to apply standard linear density wave
% theory for results interpretation. 
% This allows us to consider
%an isolated disc.

The 2D disc equations are obtained by replacing $\rho$ with the surface
mass density $\Sigma$, $p$ becomes the vertically-integrated pressure, 
and the 2D fluid velocity $\bm{v}$ is evaluated at the midplane, as are
the forces in the momentum equations. In the Poisson equation, $\rho$ is
replaced by $\Sigma\delta(z)$, where $\delta(z)$ is the
delta function.

\subsection{Disc model and initial conditions}
For the initial disc profile we adopt a modified power-law disc with
surface density given by
\begin{align}
  \Sigma(R) = \Sigma_\mathrm{ref} \left(\frac{R}{R_0}\right)^{-s}\times B(R;
  R_{1}, R_{2}, \epsilon, \delta R), 
\end{align}
where $s$ is the power-law index describing the smooth disc and
$\Sigma_\mathrm{ref}$ is a 
surface density scale chosen by specifying $Q_\mathrm{out}$,
the Keplerian Toomre parameter at $R=R_{2}$,
\begin{align}
  Q_\mathrm{out} = \left.\frac{c_s\Omega_k}{\pi G
    \Sigma}\right|_{R=R_{2}}. 
\end{align}
The bump function
$B(R)$ represents a surface density boost between
$R\in[R_{1},R_{2}]$ by a factor $\epsilon^{-1}>1$,
and $\delta R$ is the transition width between the bump and the
smooth disc. We choose 
\begin{align}
  &B(R) = f_1(R)\times f_2(R),\\
  &f_1(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 +
    \tanh\left(\frac{R-R_{1}}{\Delta_1}\right)\right]  + \epsilon,\\
  &f_2(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 -
    \tanh\left(\frac{R-R_{2}}{\Delta_2}\right)\right]  + \epsilon,
\end{align}
where $\Delta_{1,2} = \delta R \times H(R_{1,2})$. The bump function
mimicks mass built-up in $R\in[R_1,R_2]$.

@@

Our fiducial parameter values are: $s=2$, $R_{1}=R_0$, $R_{2}=2R_0$,
$\epsilon=0.1$, $\delta R=5$, $h=0.05$ and
$Q_\mathrm{out}=2$. An example of the initial surface density and the
Toomre $Q$ parameter (see\S\ref{wkb}) is shown in
Fig. \ref{initial_surf}. Since $Q>1$, the disc is stable to local
axisymmetric perturbations \citep{toomre64}. %min Q is 1.3
The transition between self-gravitating and 
non-self-gravitating portions of the disc occur smoothly across
$\sim10H$. Initially there is no vertical or radial velocity
($v_R = v_r = v_\theta = 0$). The azimuthal velocity is initialized to
satisfy centrifugal balance with pressure and gravity,
\begin{align}
  \frac{v_\phi^2}{r} = \frac{1}{\rho}\frac{\p p}{\p r} + \frac{\p
    \Phi_\mathrm{tot}}{\p r}
\end{align}
and similarly in 2D.

@@

\subsection{FARGO}
Our primary code is FARGO with self-gravity \citep{baruteau08}. This
is a popular, simple finite-difference code for 2D discs. `FARGO' refers
to its azimuthal transport algorithm, which circumvents time-step
contraint set by the inner disc boundary \citep{masset00a,masset00b}.

The 2D disc occupies
$R\in[R_\mathrm{min},R_\mathrm{max}],\,\phi\in[0,2\pi]$ and is 
divided into $(N_R,N_\phi)$ grids, logarithimically spaced in radius and
uniformly spaced in azimuth. At radial boundaries we set the
hydrodynamic variables to their initial values.

@@

The normalised surface density with azimuthal wavenumber $m$ is
\begin{align}
  \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
      m\phi)}\right]
\end{align}
where $\Sigma_{00} = \Sigma_0(t=0)$. The time evolution of the
$m^\mathrm{th}$ mode can be characterized by 
assuming $\Sigma_m\propto\exp{(-\ii \sigma t)}$, where
$\sigma = \omega + \ii\gamma$ is a complex frequency, $\omega$ being 
the real frequency and $\gamma$ is the growth rate. The total
non-axisymmetric surface density is 
\begin{align}
  \Delta\Sigma = \frac{\Sigma - \Sigma_0}{\Sigma_0}. 
\end{align}

\subsubsection{Angular momentum decomposition}
The total disc angular momentum is
\begin{align}
  J &= \int_{R_\mathrm{min}}^{R_\mathrm{max}}\int_0^{2\pi} \Sigma Rv_\phi RdRd\phi \notag\\
  &= 2\pi\int_{R_\mathrm{min}}^{R_\mathrm{max}} R\Sigma_0 v_{\phi0} R dR \notag\\ 
  &\phantom{=}+
  \sum_{m=1}^\infty2\pi\int_{R_\mathrm{min}}^{R_\mathrm{max}} 2R\real\left[\Sigma_m v_{\phi
      m}^*\right] RdR 
  %&= \sum_{m=0}^\infty \int _{R_\mathrm{min}}^{R_\mathrm{max}}  j_m(R)
  %dR 
  = \sum_{m=0}^\infty J_m. 
\end{align}
where $^*$ denotes complex conjugation. We will refer to $J_m$ as the
$m^\mathrm{th}$ component of the total angular momentum, and use it to
monitor angular momentum conservation in the numerical simulations. 
It is important to distinguish this empirical definition from the
angular momentum of linear perturbations, defined through a
conservation law, which is given in \S\ref{wkb}.

@@

\subsubsection{Three-dimensionality}
In 3D simulations we measure the importance of vertical motion with
$\Theta$, where 
\begin{align}\label{theta}
  \Theta^2 \equiv \frac{\avg{v_z^2}}{\avg{v_R^2}+\avg{v_\phi^2}}, 
\end{align}
and $\avg{\cdot}$ denotes the density-weighted average, e.g., 
\begin{align}
  \avg{v_z^2} \equiv\frac{
    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
    \rho v_z^2 dV}{
    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
    \rho dV
  },
\end{align}
and similarly for the horizontal velocities. Thus $\Theta$ is the
ratio of the average kinetic energy associated with vertical motion to
that in horizontal motion. The radial range of
integration is taken over $r\in[R_1,R_2]$ since this is where
we find the perturbations to be confined.

\section{Linear density waves}\label{wkb}
In this paper we take a fully numerical approach using non-linear
hydrodynamic simulations. However, it will be useful to employ
linear density wave theory to interpret the simulations.
We list here the main definitions and results from 2D linear theory,
taken from \cite{lin11b} and \cite{shu91}.

@@

The background torque density $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state. In a barotropic fluid, such as a strictly isothermal disc,
$\tbg$ vanishes and the angular momentum associated with
the perturbation is conserved, provided that there is no net angular
momeumtum flux and the perturbations are regular.

@@

\subsection{Local results}
In the local approximation, perturbations are assumed to vary rapidly
relative to any background gradients. The
dispersion relation for tightly-wound density 
waves of the form $\exp{[\ii(-\sigma t + m \phi + kR)]}$ in a razor-thin
disc is  
\begin{align}\label{dispersion}
  (\sigma - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
\end{align}
where $k$ is a real wavenumber such that $|kR|\gg1$ \citep{shu91}. 
Note that in this approximation, only axisymmetric perturbations with
$m=0$ can be unstable (having an imaginary $\sigma$).

Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv  
\omega/m$ of a neutral mode 
, Eq. \ref{dispersion} can be solved 
for $|k|$, 
\begin{align}\label{wavenumber}
  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
      Q^2(1-\nu^2)}\right], 
\end{align}
where 
% \begin{align}
%   k_T \equiv \frac{\kappa^2}{2\pi G \Sigma}
% \end{align}
% is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. In
Eq. \ref{wavenumber}, the upper (lower) sign correspond to short
(long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
spirals.

@@

A detailed discussion of the properties of local density waves 
is given in \cite{shu91}. An important result is that
waves interior to co-rotation ($R<R_c$) have negative angular momentum, while
waves outside co-rotation ($R>R_c$) have positive angular
momentum. The local approximation becomes less good for gobal modes,
but will nevertheless be useful for results interpretation.

### 2014-11-12 paper/paper.tex
minor updates

\maketitle
\begin{abstract} 
  We present numerical simulations of astrophysical discs with an 
  imposed temperature profile and in which a sub-portion of it is
  self-gravitating. This might represent, for example, mass  
  built-up in a circumstellar or protoplanetary
  disc at the outer edge of a `dead zone', in which there
  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
  We conduct numerical gravito-hydrodynamic experiments in 2D using
  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
  discs we observe the development of a 
  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly stationary 
  one-arm spiral pattern in the self-gravitating region of the
  disc. The development of this one-arm 
  spiral does not require the movement of  
  the central star, as found in previous studies. Instead, we find it 
  to be associated with the imposed 
  temperature gradient and the radial disc structure. 
  We apply results from existing linear theory and attribute the
  growth of this spiral to the non-conservation of angular momentum
  for non-axisymmetric linear perturbations in discs with a forced temperature
  gradient. This may lead to an unstable exchange of angular momentum between the spiral 
  and the background disc. However, because the interaction does
  not explicitly depend on self-gravity, we speculate it may also
  affect low-frequency one-arm spirals in non-self-gravitating
  discs. 
\end{abstract}

@@

% but also natural in GI.
However, spiral arms are also naturally associated with global gravitational 
instability (GI) of differentially rotating discs
\citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale
spiral arms may provide significant angular momentum transport to enable
accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
spiral structures due to GI are potentially observable with the Atacama Large 
Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}.

% when/where does it happen? formation stage/dead zones (special for ppd)
GI can be expected in the earliest stages of 
circumstellar disc formation \citep{kratter10b,inutsuka10,tsukamoto13}. %also
% outer parts  
In the case of protoplanetary discs, which have complex structure
\citep{armitage10}, GI may also occur in `dead
zones' --- sub-portions of the disc where the magneto-rotational
instability is ineffective for angular momentum transport
\citep{gammie96,turner08,landry13} and therefore has reduced mass
accretion. Dead zones may become
self-gravitating with sufficient mass built-up
\citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.

% original motivation for this work (DZGI) 
Given the wide range of GIs that may occur in  gaseous 
astrophysical discs
\citep{papaloizou89,christo92,christo93,hadley11,hadley14}, it is of 
theoretical interest to clarify what type of GI is relevant to 
partially self-gravitating discs, such as a protoplanetary disc containing a 
massive dead zone.  %This was indeed the original goal of this project. 
This work presents a first step towards this goal. We begin with
the simplest disc models which, surprisingly, still produced
unexpected results. Specifically, our seemingly simple initial disc models
consistently developed \emph{one-armed} spiral structures, despite not 
satisfying the usual requirements to generate them 
(see below).

@@

%brief history of m=1 
%stellar wobble from indiana group 
Single-arm spirals, or eccentric modes, corresponding to perturbations 
with azimuthal wavenumber $m=1$, have received interest in
the context of protoplanetary discs because of their global nature
\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. Their development is 
often associated with the unique property of $m=1$ perturbations to
move the central star \citep{shu90}, which may have observable
consequeces \citep{michael10}. 
%these are self-gravitating instabilities

@@

Our study is based on direct hydrodynamic simulations using three
different grid-based codes, both in two-dimensions (2D) and
three-dimensions (3D). We also apply results from previous linear
theory to explain their growth. We find that, when the disc
temperature is prescribed, the usual statement for the conservation of
angular momentum for linear perturbations acquires a source term
proportional to the temperature gradient, which may act to destabilise
$m=1$ perturbations. Although we work with self-gravitating discs, the
source term itself does not depend explicitly on disc self-gravity. Thus, the
destabilisation effect we describe may also operate in
non-self-gravitating discs.

This paper is organised as follows. In \S\ref{model} we list the
governing equations and describe the disc equilibria under
consideration. \S\ref{methods} describes the hydrodynamic codes
employed in this work and several diagnostic measures. We briefly
review basic results from linear theory in \S\ref{wkb} to aid our
interpretation of numerical simulations.  Our main results are presented in 
\S\ref{results2d} for 2D simulations and \S\ref{results3d} for 3D
simulations, and we further analyse them in \S\ref{discussions}. 
We summarise in \S\ref{summary}, including a discussion on
spectulations for future work.

>>>>

\maketitle
\begin{abstract} 
  We present numerical simulations of astrophysical discs with an 
  imposed temperature profile and in which a sub-portion of it is
  self-gravitating. This might represent, for example, mass  
  built-up in a circumstellar or protoplanetary
  disc at the outer edge of a `dead zone', in which there
  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
  We conduct numerical gravito-hydrodynamic experiments in 2D using
  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
  discs we observe the development of a 
  spiral mode with azimuthal wavenumber $m=1$ and co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly stationary 
  one-arm spiral pattern in the self-gravitating region of the
  disc. The development of this one-arm 
  spiral does not require the movement of  
  the central star, as found in previous studies. Instead, we find it 
  is associated with the imposed 
  temperature gradient and the radial disc structure. 
  We apply results from linear theory and attribute the
  growth of this spiral to the non-conservation of angular momentum
  for non-axisymmetric linear perturbations in discs with a forced temperature
  gradient. This may lead to an unstable exchange of angular momentum between the spiral 
  and the background disc. However, because the interaction does
  not explicitly depend on self-gravity, we suggest it may also
  affect low-frequency one-arm spirals in non-self-gravitating
  discs. 
\end{abstract}

@@

% but also natural in GI.
However, spiral arms are also associated with global gravitational 
instability (GI) of differentially rotating discs
\citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale
spiral arms can provide significant angular momentum transport
neccessary for accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
spiral structures due to GI are potentially observable with the Atacama Large 
Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}.

% when/where does it happen? formation stage/dead zones (special for ppd)
GI can be expected in the earliest stages of 
circumstellar disc formation
\citep{kratter10b,inutsuka10,tsukamoto13}, and may be possible in the
outer parts of the disc \citep{rafikov05,matzner05,kimura12}.  
% outer parts  
In the case of protoplanetary discs, which have complex structure
\citep{armitage10}, GI can also occur in `dead
zones' --- sub-portions of the disc where the magneto-rotational
instability is ineffective for angular momentum transport
\citep{gammie96,turner08,landry13} and therefore has reduced mass
accretion. Dead zones may then become
self-gravitating with sufficient mass built-up
\citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.

% original motivation for this work (DZGI) 
A wide range of GIs can occur in  gaseous 
astrophysical discs 
\citep{papaloizou89,christo92,christo93,hadley11,hadley14}. It is then
of theoretical importance to clarify what type of GI is relevant to 
partially self-gravitating discs, such as a protoplanetary disc containing a 
massive dead zone.  %This was indeed the original goal of this project. 
This work presents a first step towards this goal. Specifically, we 
study a class of seemingly simple disc models which were found to 
consistently produce \emph{one-armed} spiral structures, despite not 
satisfying the usual requirements to generate them.

@@

%brief history of m=1 
%stellar wobble from indiana group  
Single-arm spirals, or eccentric modes, corresponding to perturbations 
with azimuthal wavenumber $m=1$, have received interest in
the context of protoplanetary discs because of their global nature
\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. 
In the `SLING' mechanism proposed by \cite{shu90}, the $m=1$ spiral
instability is attributed to motion of the central star induced by the
one-arm perturbation and requires a massive disc \citep[the former may
have observable consequences, ][]{michael10}. However, our numerical
models do not have these features.

%these are self-gravitating instabilities

@@

Our study is based on direct hydrodynamic simulations using three
different grid-based codes, both in two-dimensions (2D) and
three-dimensions (3D). We also apply results from linear
theory to explain their growth. We find that, when the disc
temperature is prescribed, the usual statement for the conservation of
angular momentum for linear perturbations acquires a source term
proportional to the temperature gradient, which may act to destabilise
$m=1$ perturbations. Although we work with self-gravitating discs,
this source term does not depend explicitly on disc self-gravity. Thus, the
destabilisation effect we describe should also be applicable to
non-self-gravitating discs.

This paper is organised as follows. In \S\ref{model} we list the
governing equations and describe the disc equilibria under
consideration. \S\ref{methods} describes the hydrodynamic codes
employed in this work and several diagnostic measures. We briefly
review basic results from linear theory in \S\ref{wkb} to aid our
interpretation of numerical simulations.  Our main results are presented in 
\S\ref{results2d} for 2D simulations and \S\ref{results3d} for 3D
simulations, and we further analyse them in \S\ref{discussions}. 
We summarise in \S\ref{summary}, with a discussion on
spectulations for future work.
